\documentclass[10pt, a4paper, landscape]{extarticle}

% -----packages-----
\usepackage{multicol} % for multiple columns
\usepackage[landscape]{geometry} % for landscape
\usepackage{parskip} % remove text indentation
\usepackage{graphicx} % for scale tables
\usepackage{enumitem} % indent of lists
\usepackage{tikz} % for plots
\usepackage{hyperref} % for hyperlinks
\usepackage{amsmath} % for writing normal text on equations
\usepackage{scrlayer-scrpage} % page foot
\usepackage[compact]{titlesec} % titles spacing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% WATERMARK (NOT PERMANENT...) %%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{draftwatermark}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% -----page customization-----
\geometry{top=1cm,left=1cm,right=1cm,bottom=1cm} % margins configuration
\pagenumbering{gobble} % remove page numeration
\setlength{\parskip}{0cm} % paragraph skip length
% title spacing:
\titlespacing{\section}{0pt}{2ex}{1ex}
\titlespacing{\subsection}{0pt}{1ex}{0ex}
\titlespacing{\subsubsection}{0pt}{0.5ex}{0ex}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% WATERMARK PROPERTIES %%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\SetWatermarkText{DRAFT}
\SetWatermarkScale{3}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% -----document-----
\begin{document}

% page foot
\cfoot{\href{https://github.com/marcelomijas/econometrics-cheatsheet}{\normalfont \footnotesize Version pr0.3-en - github.com/marcelomijas/econometrics-cheatsheet}}
\setlength{\footskip}{12pt}

\begin{multicols}{3} % set columns to 3

\begin{center}
	\textbf{\LARGE \href{https://github.com/marcelomijas/econometrics-cheatsheet}{Additional Problems \\ Cheat Sheet}}
	\\ {\footnotesize By Marcelo Moreno - King Juan Carlos University}
	\\ {\footnotesize As part of the Econometrics Cheat Sheet Project}
\end{center}

\colorbox{yellow}{THIS IS A WORK IN PROGRESS}
\\ \colorbox{yellow}{NOT INTENDEND FOR GENEAL PURPOSE}

\section*{Problem of endogenous explanatory variables}
	Omitted variables provoke OLS inconsistency. Solution: proxy variables.
	\begin{itemize}[leftmargin=*]
		\item The IV method: can also solve variable errors.
		\item The TSLS.
		\begin{center}
			$y = \beta_0 + \beta_1 x + u$
		\end{center}
	\end{itemize}
	The problem of the bias of omitted variables. Solution:
	\begin{itemize}[leftmargin=*]
		\item Proxy variables.
		\item If proxy variables are not available, search for one with a relation with $x$. A $z$ variable that meet the requirements:
		\begin{center}
			$Cov(z,u) = 0$
			\\ $Cov(z,x) \neq 0$
		\end{center}
		If the requirements are meet, $z$ is an Instrumental Variable (IV).
	\end{itemize}
	\subsection*{TSLS}
		Can have multiple instrumental variables (is the IV, but with various instrumental variables at the same time). And $Cov(z,u) = 0$ can be relaxed, but there has to be a minimum of variables that satisfies it.
		\\ Can have multicollinearity problems.
		\\ \textbf{Some tests}:
		\begin{itemize}[leftmargin=*]
			\item Endogeneity tests:  is TSLS better than OLS when there are no endogenous variables? Do we really need TSLS? $\rightarrow$ Hausman test $\rightarrow H_0$: OLS is consistent (it is better to use OLS).
			\item Over-identification. An IV should meet:
			\begin{itemize}[leftmargin=*]
				\item $Corr(z,u) = 0$ (exogeneity)
				\item $Corr(z,x) \neq 0$ (relevance)
			\end{itemize}
			\item Is there too many IV? $\rightarrow$ Sagan test $\rightarrow H_0$: all IV seem ok
		\end{itemize}
	
\columnbreak

\section*{Bias of omitted variable}
	By specification error.
	\\ True model:
	\begin{center}
		$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + u$
	\end{center}
	The estimated model:
	\begin{center}
		$\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1 x_1$
	\end{center}
	If $Corr(x_1, x_2) = 0$ there is no bias.
	\\ Bias summary:
	 \\ \scalebox{0.872}{ % scale the table to fit in the column
		\begin{tabular}{| c | c | c |}
			\hline
			 & $Corr(x_1, x_2) > 0$ & $Corr(x_1, x_2) < 0$ \\
			\hline
			$\beta_2 > 0$ & $(+) bias$ & $(-) bias$ \\
			\hline
			$\beta_2 < 0$ & $(-) bias$ & $(+) bias$ \\
			\hline
		\end{tabular}
	}
	\begin{itemize}[leftmargin=*]
		\item (+) bias $\rightarrow$ $\beta_1$ will be higher than it should be (it includes the effect of $x_2$).
		\item (-) bias $\rightarrow$ $\beta_1$ will be lower than it should be (in includes the effect of $x_2$).
	\end{itemize}

\section*{Incorrect functional forms}
	Ramsey RESET test: it test the specification errors of a regression. $H_0$: the model is correctly specified.

\columnbreak

\end{multicols}

\end{document}